\begin{sloppypar}

This five-year journey has been nothing short of extraordinary. Along the way, I’ve had the privilege of learning from countless individuals who have guided, supported, and believed in me. Their encouragement has been instrumental in helping me grow into a better version of myself.

This thesis is a culmination of the values and ideas imparted to me by Dr. Aniket Alam. I am deeply grateful for his invaluable guidance throughout my journey. His insights, encouragement, and  mentorship have consistently sparked my curiosity and shaped the direction of my work. His guidance has not only helped my academic journey but also help me become the person I am today.

I would also like to extend my heartfelt thanks to all my friends at IIIT. A special mention goes to Hitesh, Shreyansh, Gaurav, Pradhuman for always being there, for listening to my rants, watching cricket matches together, and for being sources of strength and joy through all the ups and downs. The memories I’ve made with them are some of the fondest of my time here and I will always reminisce the time we spent together.

Thank you to Vansh for being an amazing lab partner and  his help with thesis writing, Radhe for being my badminton buddy, Hari for being hilarious, Srishti and Krishnakoli for being my hinges of support,  NDA for the late night games.

A big shoutout to all my CHD comrades who have made this often tumultuous five year journey not just bearable, but meaningful.

I am especially grateful to Mama, Mami, Nana, and Nani for always being there whenever I needed them, offering their wisdom and unshakable support.

Lastly, I owe the deepest gratitude to Mumma, Papa, and Ridhima for always believing in me even when I struggled to believe in myself. You have been my unwavering pillars of strength, lifting me in difficult moments, and offering your unconditional love and support. I couldn’t have done this without you.

\end{sloppypar}